---
title: "Instrumental Variable Estimation 2: Implementation in R"
author: "Instructor: Yuta Toyama"
date: "Last updated: `r Sys.Date()`"
fig_width: 6 
fig_height: 4 
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    nature:
      highlightStyle: github
      highlightLines: yes
      countIncrementalSlides: no
      ratio: '16:9'
knit: pagedown::chrome_print
---
class: title-slide-section, center, middle
name: logistics

# Introduction

---
## Introduction

- I cover three examples of instrumental variable regressions.
    1. Wage regression
    2. Demand curve
    3. Effects of Voter Turnout (Hansford and Gomez)

---
class: title-slide-section, center, middle
name: logistics
# Wage regression

---
## Example 1: Wage regression

- Use dataset "Mroz", cross-sectional labor force participation data that accompany "Introductory Econometrics" by Wooldridge. 
    - Original data from  *"The Sensitivity of an Empirical Model of Married Women's Hours of Work to Economic and Statistical Assumptions"* by Thomas Mroz published in *Econometrica* in 1987.
    - Detailed description of data: https://www.rdocumentation.org/packages/npsf/versions/0.4.2/topics/mroz 


```{r, message=FALSE}
library("foreign")

# You do not have to worry about a message "cannot read factor labels from Stata 5 files".
data <- read.dta("data/MROZ.DTA")
```

---
.pull-left-left[
- Describe data
```{r, message=FALSE, eval=FALSE}
library(stargazer)
stargazer(data,
          type = "text")
```
]

.pull-right-right[
```{r,message=FALSE, echo=FALSE}
library(stargazer)
stargazer(data,type = "text")
```
]
---
<br/><br/>
- Consider the wage regression 
$$\log(w_i) = \beta_0 + \beta_1 educ_i + \beta_2 exper_i + \beta_3 exper_i^2 + \epsilon_i$$
- We assume that $exper_i$ is exogenous but $educ_i$ is endogenous.
- As an instrument for $educ_i$, we use the years of schooling for his or her father and mother, which we call $fathereduc_i$ and $mothereduc_i$. 
- Discussion on these IVs will be later. 

---

```{r,eval=FALSE}
library("AER")
library("dplyr")
library("texreg")
library("estimatr")


# data cleaning
data %>% 
  select(lwage, educ, exper, expersq, motheduc, fatheduc) %>%
  filter( is.na(lwage) == 0 ) -> data


result_OLS <- lm_robust( lwage ~ educ + exper + expersq, data = data, se_type = "HC1")

# IV regression using fathereduc and mothereduc
result_IV <- iv_robust(lwage ~ educ + exper + expersq | 
                           fatheduc + motheduc + exper + expersq, 
                       data = data, se_type = "HC1")

# Show result
screenreg(l = list(result_OLS, result_IV),digits = 3,
          # caption = 'title',
         # custom.model.names = c("(I)", "(II)", "(III)", "(IV)", "(V)"),
          custom.coef.names = NULL,  # add a class, if you want to change the names of variables.
          include.ci = F,include.rsquared = FALSE, include.adjrs = TRUE, include.nobs = TRUE, 
          include.pvalues = FALSE, include.df = FALSE, include.rmse = FALSE,
          custom.header = list("lwage" = 1:2), # you can add header especially to indicate dependent variable 
          stars = numeric(0))

```
---
```{r,echo=FALSE,message=FALSE,warning=FALSE}
library("AER")
library("dplyr")
library("texreg")
library("estimatr")


# data cleaning
data %>% 
  select(lwage, educ, exper, expersq, motheduc, fatheduc) %>%
  filter( is.na(lwage) == 0 ) -> data


result_OLS <- lm_robust( lwage ~ educ + exper + expersq, data = data, se_type = "HC1")

# IV regression using fathereduc and mothereduc
result_IV <- iv_robust(lwage ~ educ + exper + expersq | fatheduc + motheduc + exper + expersq, data = data, se_type = "HC1")


# Show result
screenreg(l = list(result_OLS, result_IV),
          digits = 3,
          # caption = 'title',
         # custom.model.names = c("(I)", "(II)", "(III)", "(IV)", "(V)"),
          custom.coef.names = NULL,  # add a class, if you want to change the names of variables.
          include.ci = F,
          include.rsquared = FALSE, include.adjrs = TRUE, include.nobs = TRUE, 
          include.pvalues = FALSE, include.df = FALSE, include.rmse = FALSE,
          custom.header = list("lwage" = 1:2), # you can add header especially to indicate dependent variable  
          stars = numeric(0)
          )

```
---

- How about the first stage? You should always check this!!

```{r}

# First stage regression 

result_1st <- lm(educ ~ motheduc + fatheduc + exper + expersq, data = data)

# F test
linearHypothesis(result_1st, 
                 c("fatheduc = 0", "motheduc = 0" ), 
                 vcov = vcovHC, type = "HC1")

```

---
## Discussion on IV

- Labor economists have used family background variables as IVs for education. 
    - **Relevance**: OK from the first stage regression. 
    - **Independence**: A bit suspicious. Parents' education would be correlated with child's ability through quality of nurturing at an early age. 
- Still, we can see that these IVs can mitigate (though may not eliminate completely) the omitted variable bias. 
- Discussion on the validity of instruments is crucial in empirical research. 

---
class: title-slide-section, center, middle
name: logistics
# Demand curve

---
## Example 2: Estimation of the Demand for Cigaretts

- Demand model is a building block in many branches of Economics. 
- For example, health economics is concerned with the study of how health-affecting behavior of individuals is influenced by the health-care system and regulation policy. 
- Smoking is a prominent example as it is related to many illnesses and negative externalities.
- It is plausible that cigarette consumption can be reduced by taxing cigarettes more heavily. 
- Question: how much taxes must be increased to reach a certain reduction in cigarette consumption? -> Need to know **price elasticity of demand** for cigaretts.

---
<br/><br/>
- Use `CigarrettesSW` in the package `AER`.
- a panel data set that contains observations on cigarette consumption and several economic indicators for all 48 continental federal states of the U.S. from 1985 to 1995.
- What is **panel data**? The data involves both time series and cross-sectional information.
    - The variable is denoted as $y_{it}$, which indexed by individual $i$ and time $t$.
    - Cross section data $y_i$: information for a particular individual $i$ (e.g., income for person $i$).
    - Time series data $y_t$: information for a particular time period (e.g., GDP in year $y$)
    - Panel data $y_{it}$: income of person $i$ in year $t$.
- We will see more on panel data later in this course. For now, we use the panel data as just  cross-sectional data (**pooled cross-sections**)

---


```{r}
# load the data set and get an overview
data("CigarettesSW")
summary(CigarettesSW)
```

---
<br/>
- Consider the following model $$\log (Q_{it}) = \beta_0 + \beta_1 \log (P_{it}) + \beta_2 \log(income_{it}) + u_{it}$$ where 
    - $Q_{it}$ is the number of packs per capita in state $i$ in year $t$, 
    - $P_{it}$ is the after-tax average real price per pack of cigarettes, and 
    - $income_{it}$ is the real income per capita. This is demand shifter.
- As an IV for the price, we use the followings:
    - $SalesTax_{it}$: the proportion of taxes on cigarettes arising from the general sales tax.
        - Relevant as it is included in the after-tax price
        - Exogenous(indepndent) since the sales tax does not influence demand directly, but indirectly through the price.
    - $CigTax_{it}$: the cigarett-specific taxes
      
---

```{r}
CigarettesSW %>% 
  mutate( rincome = (income / population) / cpi) %>% 
  mutate( rprice  = price / cpi ) %>% 
  mutate( salestax = (taxs - tax) / cpi ) %>% 
  mutate( cigtax = tax/cpi ) -> Cigdata

```

- Let's run the regressions

```{r,eval=FALSE}
 
cig_ols <- lm_robust(log(packs) ~ log(rprice) + log(rincome) ,  data = Cigdata,se_type = "HC1")
#coeftest(cig_ols, vcov = vcovHC, type = "HC1")


cig_ivreg <- iv_robust(log(packs) ~ log(rprice) + log(rincome)  | 
                    log(rincome) +  salestax +  cigtax, data = Cigdata, se_type = "HC1")
#coeftest(cig_ivreg, vcov = vcovHC, type = "HC1")

# Show result
screenreg(l = list(cig_ols, cig_ivreg),digits = 3,
          # caption = 'title',
          custom.model.names = c("OLS", "IV"),custom.coef.names = NULL,  # add a class, if you want to change the names of variables.
          include.ci = F,include.rsquared = FALSE, include.adjrs = TRUE, include.nobs = TRUE, 
          include.pvalues = FALSE, include.df = FALSE, include.rmse = FALSE,
          custom.header = list("log(packs)" = 1:2), # you can add header especially to indicate dependent variable 
          stars = numeric(0)
          )
```

---
```{r,echo=FALSE}

cig_ols <- lm_robust(log(packs) ~ log(rprice) + log(rincome) ,  data = Cigdata,se_type = "HC1")
#coeftest(cig_ols, vcov = vcovHC, type = "HC1")


cig_ivreg <- iv_robust(log(packs) ~ log(rprice) + log(rincome)  | 
                    log(rincome) +  salestax +  cigtax, data = Cigdata, se_type = "HC1")
#coeftest(cig_ivreg, vcov = vcovHC, type = "HC1")

# Show result
screenreg(l = list(cig_ols, cig_ivreg),
          digits = 3,
          # caption = 'title',
          custom.model.names = c("OLS", "IV"),
          custom.coef.names = NULL,  # add a class, if you want to change the names of variables.
          include.ci = F,
          include.rsquared = FALSE, include.adjrs = TRUE, include.nobs = TRUE, 
          include.pvalues = FALSE, include.df = FALSE, include.rmse = FALSE,
          custom.header = list("log(packs)" = 1:2), # you can add header especially to indicate dependent variable 
          stars = numeric(0)
          )
```

---
<br/>
- The first stage regression
```{r,eval=FALSE}

# First stage regression 

result_1st <- lm(log(rprice) ~ log(rincome) + log(rincome) + salestax + cigtax , data= Cigdata)

# F test
linearHypothesis(result_1st, 
                 c("salestax = 0", "cigtax = 0" ), 
                 vcov = vcovHC, type = "HC1")

```

---
<br/><br/>
```{r, echo=FALSE}

# First stage regression 

result_1st <- lm(log(rprice) ~ log(rincome) + log(rincome) + salestax + cigtax , data= Cigdata)

# F test
linearHypothesis(result_1st, 
                 c("salestax = 0", "cigtax = 0" ), 
                 vcov = vcovHC, type = "HC1")

```

---
class: title-slide-section, center, middle
name: logistics

# Voting

---
## Example 3: Effects of Turnout on Partisan Voting

- THOMAS G. HANSFORD and BRAD T. GOMEZ "Estimating the Electoral Effects of Voter Turnout" The American Political Science Review Vol. 104, No. 2 (May 2010), pp. 268-288
    - Link: https://www.cambridge.org/core/journals/american-political-science-review/article/estimating-the-electoral-effects-of-voter-turnout/8A880C28E79BE770A5CA1A9BB6CF933C
- Here, we will see a simplified version of their analysis. 
- The dataset is [here](HansfordGomez_Data.csv)

---

```{r, eval=FALSE}
library(readr)

HGdata <- read_csv("data/HansfordGomez_Data.csv")

stargazer::stargazer(as.data.frame(HGdata) %>% select(-starts_with("Yr")),type="text")

```
---


```{r, echo=FALSE,message=FALSE,warning=FALSE}
library(readr)

HGdata <- read_csv("data/HansfordGomez_Data.csv")

stargazer::stargazer(as.data.frame(HGdata) %>% select(-starts_with("Yr"),-(dph_StateVAP:State_DVS_lag2)),type="text")

```
---
```{r, echo=FALSE,message=FALSE,warning=FALSE}

stargazer::stargazer(as.data.frame(HGdata) %>% select(dph_StateVAP:State_DVS_lag2),type="text")

```
---

- Data description: 

| Name | Description|
|---|-------------------------------------------------------------------|
| Year | Election Year|
| FIPS_County | FIPS County Code|
| Turnout | Turnout as Pcnt VAP|
| Closing2 | Days between registration closing date and election|
| Literacy | Literacy Test|
| PollTax | Poll Tax|
| Motor | Motor Voter|
| GubElection | Gubernatorial Election in State|
| SenElection | U.S. Senate Election in State|
| GOP_Inc | Republican Incumbent|


---


| Name | Description|
|---|-------------------------------------------------------------------|
| Yr52 | 1952 Dummy|
| Yr56 | 1956 Dummy|
| Yr60 | 1960 Dummy|
| Yr64 | 1964 Dummy|
| Yr68 | 1968 Dummy|
| Yr72 | 1972 Dummy|
| Yr76 | 1976 Dummy|
| Yr80 | 1980 Dummy|
| Yr84 | 1984 Dummy|
| Yr88 | 1988 Dummy|
| Yr92 | 1992 Dummy|
| Yr96 | 1996 Dummy|
| Yr2000 | 2000 Dummy|

---

<br/><br/>

| Name | Description|
|---|-------------------------------------------------------------------|
| DNormPrcp_KRIG | Election day rainfall - differenced from normal rain for the day|
| GOPIT | Turnout x Republican Incumbent|
| DemVoteShare2_3MA | Partisan composition measure = 3 election moving avg. of Dem Vote Share|
| DemVoteShare2 | Democratic Pres Candidate's Vote Share|
| RainGOPI | Rainfall measure x Republican Incumbent|
| TO_DVS23MA | Turnout x Partisan Composition measure|
| Rain_DVS23MA | Rainfall measure x Partisan composition measure|
| dph | =1 if home state of Dem pres candidate|
| dvph | =1 if home state of Dem vice pres candidate|

---

<br/><br/>

| Name | Description|
|---|-------------------------------------------------------------------|
| rph | =1 if home state of Rep pres candidate|
| rvph | =1 if home state of Rep vice pres candidate|
| state_del | avg common space score for the House delegation|
| dph_StateVAP | = dph*State voting age population|
| dvph_StateVAP | = dvph*State voting age population|
| rph_StateVAP | = rph*State voting age population|
| rvph_StateVAP | = rvph*State voting age population|
| State_DVS_lag | State-wide Dem vote share, lagged one election|
| State_DVS_lag2 | State_DVS_lag squared|

---
- Consider the following regression $$DemoShare_{it} = \beta_0 + \beta_1 Turnout_{it} + u_t + u_{it}$$ where 
    - $DemoShare_{it}$ : Two-party vote share for Democrat candidate in county $i$ in the presidential election in year $t$
    - $Turnout_{it}$ : Turnout rate in county $i$ in the presidential election in year $t$
    - $u_t$ : **Year fixed effects**. Time dummies for each presidential election year
- As an IV, we use the rainfall measure denoted by `DNormPrcp_KRIG`

---

```{r, cache=TRUE,eval=FALSE}
# You can do this, but it is tedious.
hg_ols <- lm_robust( DemVoteShare2 ~ Turnout + Yr52 + Yr56 + Yr60 + Yr64 + Yr68 + Yr72 + Yr76 + Yr80 
                     + Yr84 + Yr88 + Yr92 + Yr96 + Yr2000,  data = HGdata, se_type="HC1")
#coeftest(hg_ols, vcov = vcovHC, type = "HC1")

# By using "factor(Year)" as an explanatory variable, the regression automatically incorporates the dummies for each value.
hg_ols <- lm_robust( DemVoteShare2 ~ Turnout + factor(Year) , data = HGdata, se_type="HC1")
#coeftest(hg_ols, vcov = vcovHC, type = "HC1")

# Iv regression
hg_ivreg <- iv_robust( DemVoteShare2 ~ Turnout + factor(Year) | 
                    factor(Year) + DNormPrcp_KRIG, data = HGdata, se_type="HC1")
#coeftest(hg_ivreg, vcov = vcovHC, type = "HC1")

# Show result
screenreg(l = list(hg_ols, hg_ivreg),
          digits = 3,
          # caption = 'title',
          custom.model.names = c("OLS", "IV"),
          custom.coef.names = NULL,  # add a class, if you want to change the names of variables.
          include.ci = F,
          include.rsquared = FALSE, include.adjrs = TRUE, include.nobs = TRUE, 
          include.pvalues = FALSE, include.df = FALSE, include.rmse = FALSE,
          custom.header = list("DemVoteShare2" = 1:2), # you can add header especially to indicate dependent variable 
          stars = numeric(0)
          )
```

---
```{r, cache=TRUE,echo=FALSE}

# By using "factor(Year)" as an explanatory variable, the regression automatically incorporates the dummies for each value.
hg_ols <- lm_robust( DemVoteShare2 ~ Turnout + factor(Year) , data = HGdata, se_type="HC1")
#coeftest(hg_ols, vcov = vcovHC, type = "HC1")

# Iv regression
hg_ivreg <- iv_robust( DemVoteShare2 ~ Turnout + factor(Year) | 
                    factor(Year) + DNormPrcp_KRIG, data = HGdata, se_type="HC1")
#coeftest(hg_ivreg, vcov = vcovHC, type = "HC1")

# Show result
screenreg(l = list(hg_ols, hg_ivreg),
          digits = 3,
          # caption = 'title',
          custom.model.names = c("OLS", "IV"),
          custom.coef.names = NULL,  # add a class, if you want to change the names of variables.
          include.ci = F,
          include.rsquared = FALSE, include.adjrs = TRUE, include.nobs = TRUE, 
          include.pvalues = FALSE, include.df = FALSE, include.rmse = FALSE,
          custom.header = list("DemVoteShare2" = 1:2), # you can add header especially to indicate dependent variable 
          stars = numeric(0),
          omit.coef = "factor"
          )
```
---
```{r, cache=TRUE,eval=FALSE}
# First stage regression 
hg_1st <- lm(Turnout ~  factor(Year) + DNormPrcp_KRIG, data= HGdata)

# F test
linearHypothesis(hg_1st, 
                 c("DNormPrcp_KRIG = 0" ), 
                 vcov = vcovHC, type = "HC1")

```
---
```{r, cache=TRUE,echo=FALSE}
# First stage regression 
hg_1st <- lm(Turnout ~  factor(Year) + DNormPrcp_KRIG, data= HGdata)

# F test
linearHypothesis(hg_1st, 
                 c("DNormPrcp_KRIG = 0" ), 
                 vcov = vcovHC, type = "HC1")

```

