---
title: "Regression 2: Implementation in R"
author: "Instructor: Yuta Toyama"
date: "Last updated: `r Sys.Date()`"
fig_width: 6 
fig_height: 4 
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    nature:
      countIncrementalSlides: no
      highlightLines: yes
      highlightStyle: github
      ratio: '16:9'
knit: pagedown::chrome_print
---

class: title-slide-section, center, middle
name: logistics
# Introduction

---
## Acknowledgement
This note is based on  "Introduction to Econometrics with R". https://www.econometrics-with-r.org/index.html

---
## Preliminary: packages

- We use the following packages:
    - `AER` : 
    - `dplyr` : data manipulation
    - `stargazer` : output of regression results

``` {r, warning = FALSE, message=FALSE}

# Install package if you have not done so 
# install.packages("AER")
# install.packages("dplyr")
# install.packages("stargazer")
# install.packages("texreg")
# install.packages("estimatr")

# load packages
library("AER")
library("dplyr")
library("stargazer")
library("texreg")
library("estimatr")
```

---
## Empirical setting: Data from California School

- Question: How does the student-teacher ratio affects test scores?
- We use data from California school, which is included in `AER` package. 
    - See here for the details: https://www.rdocumentation.org/packages/AER/versions/1.2-6/topics/CASchools

```{r}
# load the the data set in the workspace
data(CASchools)
```

---

- Use `class()` function to see `CASchools` is `data.frame` object.
```{r}
class(CASchools)
```

- We take 2 steps for the analysis.
    - Step 1: Look at data (descriptive analysis) 
    - Step 2: Run regression

---
class: title-slide-section, center, middle
name: logistics
# Step 1: Descriptive analysis
---
### Descriptive analysis

- It is always important to grasp your data before running regression. 
- `head()` function give you a first overview of the data.

```{r}
head(CASchools)

```

- Alternatively, you can use `View()` to see the entire dataset in browser window.

---
## Create variables

- Create several variables that are needed for the analysis.
- We use `dplyr` for this purpose.

```{r}
CASchools %>% 
  mutate( STR = students / teachers ) %>% 
  mutate( score = (read + math) / 2 ) -> CASchools 

```

---
## Descriptive statistics


- There are several ways to show descriptive statistics
- The standard one is to use `summary()` function

```{r,eval=FALSE}
summary(CASchools)
```

---

```{r,echo=FALSE}
summary(CASchools)
```

---
- This returns the desriptive statistics for all the variables in dataframe.
- You can combine this with `dplyr::select` 

```{r}
CASchools %>% 
  select(STR, score) %>%
  summary()

```

---
- You can do a bit lengthly thing manually like this.

```{r}
# compute sample averages of STR and score
avg_STR <- mean(CASchools$STR) 
avg_score <- mean(CASchools$score)

# compute sample standard deviations of STR and score
sd_STR <- sd(CASchools$STR) 
sd_score <- sd(CASchools$score)

# set up a vector of percentiles and compute the quantiles 
quantiles <- c(0.10, 0.25, 0.4, 0.5, 0.6, 0.75, 0.9)
quant_STR <- quantile(CASchools$STR, quantiles)
quant_score <- quantile(CASchools$score, quantiles)

# gather everything in a data.frame 
DistributionSummary <- data.frame(Average = c(avg_STR, avg_score), 
                                  StandardDeviation = c(sd_STR, sd_score), 
                                  quantile = rbind(quant_STR, quant_score))
```

---
```{r}
DistributionSummary
```
---
- My personal favorite is to use `stargazer` function.

```{r}

stargazer(CASchools, type = "text")

```

---
- You can choose summary statistics you want to report.

```{r}

CASchools %>% 
  stargazer( type = "text", summary.stat = c("n", "p75", "sd") ) 

```
---
<br/><br/>
- See https://www.jakeruss.com/cheatsheets/stargazer/#the-default-summary-statistics-table for the details.
- `stargazer` can be also used to report regression results.
- But, we will use `texreg` instead.

---
## Scatter plot

- Let's see how test score and student-teacher-ratio is correlated. 

```{r, eval=FALSE}

plot(score ~ STR, 
     data = CASchools,
     main = "Scatterplot of TestScore and STR", 
     xlab = "STR (X)",
     ylab = "Test Score (Y)")

```

---
.center[
```{r, fig.height=8, fig.width=12, echo=FALSE}

plot(score ~ STR, 
     data = CASchools,
     main = "Scatterplot of TestScore and STR", 
     xlab = "STR (X)",
     ylab = "Test Score (Y)")

```
]

---
<br/><br/>
- Use `cor()` to compute the correlation between two numeric vectors.
```{r}
cor(CASchools$STR, CASchools$score)
``` 
---
class: title-slide-section, center, middle
name: logistics
# Step 2: Run regression
---
## Simple linear regression

- We use `lm()` function to run linear regression

- First, consider the simple linear regression 
$$score_i = \beta_0 + \beta_1 size_i + \epsilon_i$$
where $size_i$ is the class size (student-teacher-ratio).
    - From now on we call student-teacher-ratio (STR) class size.
    
---
- To run this regression, we use `lm`

```{r}
# First, we rename the variable `STR`
CASchools %>% 
  dplyr::rename( size = STR) -> CASchools

# Run regression and save results in the varaiable `model1_summary`
model1_summary <- lm( score ~ size, data = CASchools)

# See the results
summary(model1_summary)

```

---
- Interpretations
    - An increase of one student per teacher leads to 2.2 point decrease in test scores. 
    - p value is very small. The effect of the class size on test score is significant.
        - Note: Be careful. These standard errors are NOT heteroskedasiticity robust. We will come back to this point soon.
    - $R^2 = 0.051$, implying that 5.1% of the variance of the dependent variable is explained by the model.
    
- You can add more variable in the regression (will see this soon)

---
## Robust standard error with `lm_robust`

- We use `lm_robust()` in `estimatr` package to run regression with robust standard error.
```{r}

model1_robust <- lm_robust( score ~ size, data = CASchools, se_type = "HC1")

summary(model1_robust)
```

- Notice that robust standard errors are larger than the one we obtained from `lm`!

---
## Report by texreg

- `texreg` is useful to show the regression result. 
    - `screenreg` function shows the table on R markdown.
    - You can use `htmlreg` (`texreg`) to get html (latex) format.
- `stargazer` is also used to show regression results, however it does not follow `lm_robust`.
```{r, eval=FALSE}
# Create output by `screenreg` function.
screenreg(l=list(model1_summary, model1_robust),
          digits = 3,
          # caption = 'title',
          custom.model.names = c("model1", "model1 Robust"),
          custom.coef.names = NULL,  # add a class, if you want to change the names of variables.
          include.ci = F,
          include.rsquared = FALSE, include.adjrs = TRUE, include.nobs = TRUE, 
          include.pvalues = FALSE, include.df = FALSE, include.rmse = FALSE,
          custom.header = list("score" = 1:2), # you can add header especially to indicate dependent variable 
          stars = numeric(0) # to delete star expression
          )
```


---


```{r, echo = FALSE }
# Create output by `screenreg` function.
screenreg(l=list(model1_summary, model1_robust),
          digits = 3,
          # caption = 'title',
          custom.model.names = c("model1", "model1 Robust"),
          custom.coef.names = NULL,  # add a class, if you want to change the names of variables.
          include.ci = F,
          include.rsquared = FALSE, include.adjrs = TRUE, include.nobs = TRUE, 
          include.pvalues = FALSE, include.df = FALSE, include.rmse = FALSE,
          custom.header = list("score" = 1:2), # you can add header especially to indicate dependent variable 
          stars = numeric(0) # to delete star expression
          )
```




---
## Full results

Taken from https://www.econometrics-with-r.org/7-6-analysis-of-the-test-score-data-set.html

```{r, eval=FALSE}

# estimate different model specifications
spec1 <- lm_robust(score ~ size, data = CASchools, se_type = "HC1")
spec2 <- lm_robust(score ~ size + english, data = CASchools, se_type = "HC1")
spec3 <- lm_robust(score ~ size + english + lunch, data = CASchools, se_type = "HC1")
spec4 <- lm_robust(score ~ size + english + calworks, data = CASchools, se_type = "HC1")
spec5 <- lm_robust(score ~ size + english + lunch + calworks, data = CASchools, se_type = "HC1")

# generate a table using texreg
screenreg(l = list(spec1, spec2, spec3, spec4, spec5),
          digits = 3,
          # caption = 'title',
          custom.model.names = c("(I)", "(II)", "(III)", "(IV)", "(V)"),
          custom.coef.names = NULL,  # add a class, if you want to change the names of variables.
          include.ci = F,
          include.rsquared = FALSE, include.adjrs = TRUE, include.nobs = TRUE, 
          include.pvalues = FALSE, include.df = FALSE, include.rmse = FALSE,
          custom.header = list("score" = 1:2), # you can add header especially to indicate dependent variable 
          stars = numeric(0) # to delete star expression
          )
```
---
```{r, echo=FALSE}

# estimate different model specifications
spec1 <- lm_robust(score ~ size, data = CASchools, se_type = "HC1")
spec2 <- lm_robust(score ~ size + english, data = CASchools, se_type = "HC1")
spec3 <- lm_robust(score ~ size + english + lunch, data = CASchools, se_type = "HC1")
spec4 <- lm_robust(score ~ size + english + calworks, data = CASchools, se_type = "HC1")
spec5 <- lm_robust(score ~ size + english + lunch + calworks, data = CASchools, se_type = "HC1")

# generate a table using texreg
screenreg(l = list(spec1, spec2, spec3, spec4, spec5),
          digits = 3,
          # caption = 'title',
          custom.model.names = c("(I)", "(II)", "(III)", "(IV)", "(V)"),
          custom.coef.names = NULL,  # add a class, if you want to change the names of variables.
          include.ci = F,
          include.rsquared = FALSE, include.adjrs = TRUE, include.nobs = TRUE, 
          include.pvalues = FALSE, include.df = FALSE, include.rmse = FALSE,
          custom.header = list("score" = 1:2), # you can add header especially to indicate dependent variable 
          stars = numeric(0) # to delete star expression
          )
```
- The coefficient on the class size decreases as we add more explantory variables. Can you explain why? (Hint: omitted variable bias)


